Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-13 paper.tex

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it easy to run simple
simulations, but as model complexity and the number of parameters increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from a distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and
a number of packages have been developed
to allow simulations to be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

@@

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated (see Fig.~\ref{fig-mutated-trees}),
with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
The \texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.
We also provide an \ms\ compatible
command line interface to support existing workflows.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination
stochastic process as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, usage of the term
has shifted away from its original definition as a stochastic process,
to being interpreted as a representation of a particular genetic ancestry
as a graph, without necessarily following the specific details of the Griffiths
formulation~\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.
Under the latter interpretation,
the tree sequence encoding of genetic ancestry (described above)
clearly \emph{is} an ARG: the nodes and edges
define a graph in which edges are annotated with the
set of disjoint genomic intervals
through which ancestry flows.

>>>>

The majority of simulation packages are controlled either through
a command line interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a text-based input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or a mixture of both.
Command line interfaces make it very easy to run simple
simulations, but as model complexity and the number of parameters specified increase,
they become difficult to understand and
error-prone~\citep{ragsdale2020lessons,gower2021demes}.
Specifying parameters through a text file alleviates this problem to a degree,
but lacks flexibility, for example, when running simulations with parameters
drawn from some distribution. In practice, for any reproducible simulation
project users will write a script
to generate the required command lines or input parameter files,
invoke the simulation engine, and process the results in some way.
This process is cumbersome and labour intensive, and
a number of packages have been developed
to allow simulations to be run directly in a high-level
scripting language~\citep{staab2016coala,parobek2017skelesim,gladstein2018simprily}.

@@

A major change for the \msprime\ 1.0 release is the introduction of a new set of APIs,
designed in part to avoid sources of error (see the Demography section) but
also to provide more appropriate defaults while keeping compatibility with
existing code. In the new APIs, ancestry and mutation simulation are fully
separated, with the \texttt{sim\_ancestry} and \texttt{sim\_mutations}
functions replacing the legacy \texttt{simulate} function. Among other changes,
the new APIs default to discrete genome coordinates and finite sites mutations,
making the default settings more realistic and resolving a major source of confusion and error.
The previous APIs are fully
supported and tested, and will be maintained for the foreseeable future.
The \texttt{msp} program has been extended to include new commands
for simulating ancestry and mutations separately. A particularly useful
feature is the ability to specify demographic models in
Demes format~\citep{gower2021demes} from the command line,
making simulation of complex demographies straightforward.
We also provide an \ms\ compatible
command line interface to support existing workflows.

@@

The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral}
as a representation of the coalescent with recombination
stochastic process as a graph.
This formulation is complementary to Hudson's earlier
work~\citep{hudson1983properties}, and substantially increased our theoretical
understanding of recombination. In Griffiths' ARG formulation,
a realisation of the coalescent with recombination is a graph in which
vertices represent common ancestor or recombination events, and edges represent
lineages. There is the ``big'' ARG, in which we track lineages arising out of
recombinations regardless of whether they carry ancestral
material~\citep{ethier1990two},
% and
% consequently wait an exponentially large time until all sampled lineages
% coalesce into a single ancestor~\citep{ethier1990two},
and the ``little'' ARG in which we only track genetic ancestors.
Over time, the interpretation of the term
has shifted away from the original definition as a stochastic
process, to a representation of a particular genetic ancestry
using common ancestor and recombination
events~\citep[e.g.][]{minichiello2006mapping},
to being synonymous with the genetic
ancestry of a sample, without any specific details of how genetic
material disperses~\citep{mathieson2020ancestry}.
Under the latter interpretation,
the tree sequence encoding of genetic ancestry (described above)
clearly \emph{is} an ARG: the nodes and edges
define a graph in which edges are annotated with the
set of disjoint genomic intervals
through which ancestry flows.
